Examining the impact Overturn of Roe v. Wade: Banning of Abortion on Infant Mortality Rates in the United States Using a Difference-in-Differences Approach

### 2024-11-28 paper/ExaminingtheimpactOverturnofRoev.Wade:BanningofAbortiononInfantMortalityRatesintheUnitedStatesUsingaDifference-in-DifferencesApproach.qmd
update paper name

We adopt a Bayesian framework to enhance the DID approach by incorporating prior information and quantifying uncertainty probabilistically. Using the rstanarm package, we apply weakly informative priors, including a Normal prior ($\mu=0$, $\sigma=2.5$) for fixed effects and an Exponential prior (rate = 1) for the auxiliary parameter, with autoscale=TRUE to dynamically adjust priors based on the data scale. This improves convergence and ensures priors align with predictor ranges without dominating the data. The Bayesian framework supports plausible parameter centering, captures unobserved heterogeneity through random effects, and ensures robust estimates within the DID framework, assuming parallel trends between groups.

@@

The model highlights key demographic and policy-driven impacts on infant mortality rates. The interaction between abortion bans and the post-injunction period is linked to increased mortality, while states enforcing abortion bans show consistently higher rates. Younger mothers and Black mothers face disproportionately higher risks, with limited access to healthcare and systemic inequities compounding their vulnerabilities, emphasizing the essential need for targeted health and social interventions. Increasing maternal age shows a protective effect, likely reflecting greater access to resources, improved prenatal care, and overall maternal stability, correlating with markedly lower infant mortality rates.These findings underscore the interplay of policy and demographics in shaping infant health outcomes, informing the need for evidence-based policies and healthcare strategies.

@@

The analysis shows distinct patterns in infant mortality rates across
various policy and demographic factors. In states where abortion
remained legal, the post-injunction period (after_injunction1) is linked
to reductions in infant mortality rates by 0.525 and 0.940 deaths per
1,000 live births in Models 1 and 2, respectively. However, states with
abortion bans (abortion_illegal1) exhibit significantly higher infant
mortality rates, with increases of 3.365 deaths in Model 1 and 1.400
deaths in Model 2. The DID interaction term (after_injunction1 ×
abortion_illegal1) shows that abortion bans during the post-injunction
period results in an additional increase of 0.111 deaths in Model 1 and
0.280 deaths in Model 2. Demographic factors also play a essential role,
with older maternal age (25–34 years) reducing infant mortality rates,
while younger mothers (15–19 years) face elevated risks. Black mothers
experience disproportionately higher infant mortality rates, with an
increase of 6.660 deaths per 1,000 live births compared to Asian
mothers. These findings collectively emphasize the compounded effects of
policy and demographic disparities on infant health outcomes.

@@

Additional variables were created to facilitate analysis: `after_injunction`, indicating whether the data corresponded to the period after June 2022 (post-Dobbs decision), and `abortion_illegal`, denoting whether the state had restrictive abortion laws. Death rate reliability was assessed, and unreliable rates were flagged for transparency. The death rate column was cleaned to remove annotations and converted to numeric format for analysis. A date variable was created by combining the year and month columns, providing a timeline for analysis. Finally, a date variable was created by combining the year and month variables. The cleaned dataset was saved in CSV and Parquet formats to ensure compatibility with analytical workflows.

@@

In @fig-ppcheckandposteriorvsprior-3 and @fig-ppcheckandposteriorvsprior-4 we compare the posterior with the prior. The posterior vs. prior plots for Model 1 and Model 2 illustrate the role of data in informing the posterior estimates of the model parameters. In Model 1 @fig-ppcheckandposteriorvsprior-3, which focuses on policy-related predictors (e.g., abortion restrictions), the data significantly shift the posterior distributions for key variables, demonstrating their explanatory power. Model 2 @fig-ppcheckandposteriorvsprior-4, incorporating additional demographic factors such as maternal age and race, shows a more refined and concentrated posterior distribution, emphasizing the added predictive utility of demographic variables. The comparison indicates that Model 2 provides better-informed parameter estimates, supporting the hypothesis that demographic factors play a essential role in understanding infant mortality in the context of restrictive abortion laws.

@@

plot(
  first_model,
  "areas"
)
plot(
  second_model,
  "areas"
)
```
Credibility intervals are the Bayesian equivalent of confidence intervals, offering understanding into the range of probable values for a parameter @tellingstories. Using Bayesian estimation, we derive a posterior distribution for each model coefficient. @fig-area-1 presents the 95% credibility intervals for coefficients in Model 1, while @fig-area-2 illustrates the same for Model 2. These intervals represent the range within which 95% of the posterior probability mass falls, offering a robust understanding of parameter uncertainty and the influence of predictors under the Bayesian framework @tellingstories.

@@

Our study examines the impact of abortion bans on infant mortality rates in the U.S. following the Dobbs decision, combining observational data and surveys. Observational data reveal trends and associations, but often struggle with confounding factors and causality. Surveys address these gaps by capturing qualitative understanding into individual experiences and contextual factors. This mixed-methods approach integrates statistical rigor with detailed personal accounts, offering a detailed framework to understand how restrictive abortion policies affect maternal health and infant outcomes, and informing evidence-based policy recommendations.

@@

The objective of this study is to investigate the psychological, social,
and systemic impacts of restrictive abortion laws on maternal health and infant outcomes. By focusing on mothers unable to access abortion services who experienced the death of their infant, the study aims to document lived experiences, explore the role of maternal stress and healthcare barriers in adverse outcomes. Understanding how maternal mental health directly affects prenatal care decisions, access to healthcare, and overall maternal behavior is essential to addressing the systemic factors contributing to higher infant mortality rates. The findings will guide the development of evidence-based strategies to improve maternal mental health, enhance healthcare access and shape policy responses to restrictive abortion laws.

@@

We aim to survey 20 respondents who meet our defined sample criteria: mothers unable to access abortion services due to restrictive laws, who experienced the death of their infant within the first year of life, and are reachable through Abortion Rights Advocacy Groups. Participation will be voluntary, and respondents must be willing to answer the survey questions, ensuring the quality and depth of the data collected.

>>>>

We adopt a Bayesian framework to enhance the DID approach by incorporating prior information and quantifying uncertainty probabilistically. Using the `rstanarm` package, we apply weakly informative priors, including a Normal prior ($\mu=0$, $\sigma=2.5$) for fixed effects and an Exponential prior (rate = 1) for the auxiliary parameter, with autoscale=TRUE to dynamically adjust priors based on the data scale. This improves convergence and ensures priors align with predictor ranges without dominating the data. The Bayesian framework supports plausible parameter centering, captures unobserved heterogeneity through random effects, and ensures robust estimates within the DID framework, assuming parallel trends between groups.

@@

The model highlights key demographic and policy-driven impacts on infant mortality rates. The interaction between abortion bans and the post-injunction period is linked to increased mortality, while states enforcing abortion bans show consistently higher rates. Younger mothers and Black mothers face disproportionately higher risks, with limited access to healthcare and systemic inequities compounding their vulnerabilities, emphasizing the essential need for targeted health and social interventions. Increasing maternal age shows a protective effect, likely reflecting greater access to resources, improved prenatal care, and overall maternal stability, correlating with markedly lower infant mortality rates. These findings underscore the interplay of policy and demographics in shaping infant health outcomes, informing the need for evidence-based policies and healthcare strategies.

@@

The analysis shows distinct patterns in infant mortality rates across
various policy and demographic factors. In states where abortion
remained legal, the post-injunction period (after_injunction1) is linked
to reductions in infant mortality rates by 0.525 and 0.940 deaths per
1,000 live births in Models 1 and 2, respectively. However, states with
abortion bans (abortion_illegal1) exhibit significantly higher infant
mortality rates, with increases of 3.365 deaths in Model 1 and 1.400
deaths in Model 2. The DID interaction term (after_injunction1 ×
abortion_illegal1) shows that abortion bans during the post-injunction
period results in an additional increase of 0.111 deaths in Model 1 and
0.280 deaths in Model 2. Demographic factors also play an essential role,
with older maternal age (25–34 years) reducing infant mortality rates,
while younger mothers (15–19 years) face elevated risks. Black mothers
experience disproportionately higher infant mortality rates, with an
increase of 6.660 deaths per 1,000 live births compared to Asian
mothers. These findings collectively emphasize the compounded effects of
policy and demographic disparities on infant health outcomes.

@@

Additional variables were created to facilitate analysis: `after_injunction`, indicating whether the data correspond to the period after June 2022 (post-Dobbs decision) and `abortion_illegal`, denoting whether the state had restrictive abortion laws. Death rate reliability was assessed, and unreliable rates were flagged for transparency. The death rate column was cleaned to remove annotations and converted to numeric format for analysis. A date variable was created by combining the year and month columns, providing a timeline for analysis. Finally, a date variable was created by combining the year and month variables. The cleaned dataset was saved in CSV and Parquet formats to ensure compatibility with analytical workflows.

@@

In @fig-ppcheckandposteriorvsprior-3 and @fig-ppcheckandposteriorvsprior-4 we compare the posterior with the prior. The posterior vs. prior plots for Model 1 and Model 2 illustrate the role of data in informing the posterior estimates of the model parameters. In Model 1 @fig-ppcheckandposteriorvsprior-3, which focuses on policy-related predictors (e.g., abortion restrictions), the data significantly shift the posterior distributions for key variables, demonstrating their explanatory power. Model 2 @fig-ppcheckandposteriorvsprior-4, incorporating additional demographic factors such as maternal age and race, shows a more refined and concentrated posterior distribution, emphasizing the added predictive utility of demographic variables. The comparison indicates that Model 2 provides better-informed parameter estimates, supporting the hypothesis that demographic factors play an essential role in understanding infant mortality in the context of restrictive abortion laws.

@@

plot(
  first_model,
  "areas"
)
plot(
  second_model,
  "areas"
)
```
Credibility intervals are the Bayesian equivalent of confidence intervals, offering an understanding of the range of probable values for a parameter @tellingstories. Using Bayesian estimation, we derive a posterior distribution for each model coefficient. @fig-area-1 presents the 95% credibility intervals for coefficients in Model 1, while @fig-area-2 illustrates the same for Model 2. These intervals represent the range within which 95% of the posterior probability mass falls, offering a robust understanding of parameter uncertainty and the influence of predictors under the Bayesian framework @tellingstories.

@@

Our study examines the impact of abortion bans on infant mortality rates in the U.S. following the Dobbs decision, combining observational data and surveys. Observational data reveal trends and associations but often struggle with confounding factors and causality. Surveys address these gaps by capturing a qualitative understanding of individual experiences and contextual factors. This mixed-methods approach integrates statistical rigor with detailed personal accounts, offering a detailed framework to understand how restrictive abortion policies affect maternal health and infant outcomes, and informing evidence-based policy recommendations.

@@

The objective of this study is to investigate the psychological, social,
and systemic impacts of restrictive abortion laws on maternal health and infant outcomes. By focusing on mothers unable to access abortion services who experienced the death of their infant, the study aims to document lived experiences, and explore the role of maternal stress and healthcare barriers in adverse outcomes. Understanding how maternal mental health directly affects prenatal care decisions, access to healthcare, and overall maternal behavior is essential to addressing the systemic factors contributing to higher infant mortality rates. The findings will guide the development of evidence-based strategies to improve maternal mental health, enhance healthcare access, and shape policy responses to restrictive abortion laws.

@@

We aim to survey 20 respondents who meet our defined sample criteria: mothers unable to access abortion services due to restrictive laws, who experienced the death of their infant within the first year of life, and who are reachable through Abortion Rights Advocacy Groups. Participation will be voluntary, and respondents must be willing to answer the survey questions, ensuring the quality and depth of the data collected.
